The Concurrent Calculi Formalisation Benchmark

### 2024-02-21 short-paper/short-paper.tex
Minor edits.

While results about concurrent
formalisms have already been mechanised (as we will discuss further
below), our experience is that choosing appropriate mechanisation
techniques and tools remains a significant challenge and that the
trade-offs are not well understood.  This often leads researchers
toward a trial-and-error approach, resulting in sub-optimal solutions,
wasted mechanisation effort, and techniques and results that are hard
to reuse. For example, the authors in~\cite{Cruz-Filipe2021b} note that the high-level parts of
mechanised proofs closely resemble the informal ones, while the
main challenge lies in getting the infra-structure right.

@@

\subsection{Linearity}

% \cite{Brady2017} notes that using algebraic effects is unwieldy when there are many communication channels and suggests an experimental approach using dependent uniqueness types.
% \cite{Zalakain2019} notes that systems lacking linear types must simulate linearity, and that using a linearity predicate complicates congruences.
% They had to limit themselves to an asymmetric structural precongruence to be able to prove subject reduction.

Linear typing systems allow  tracking the use of resources in a program. In
the case of typed process calculi (in particular, session types), linearity is
widely used for checking if and how a channel is used to send or receive values.
This leads to substructural typing~\cite[Chapter 1]{Pierce2004-oq} with
resulting mechanisation difficulties --- \eg deciding how to \emph{split the
typing context} in a parallel composition.

The goal of our challenge on linear reasoning is proving a type safety theorem
for a process calculus with session types, by combining as usual subject
reduction with absence of errors.  We have chosen  only to model linear (as
opposed to shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where a restriction $(\nu ab)\ldots$ binds two dual names $a$
and $b$ as opposite endpoints of a same channel --- and their duality is
reflected in the type system. We model a simple notion of error: a well-typed
processes must never use dual channel endpoints in a non-dual way (\eg by
performing concurrent send/receive operations on a same endpoint, or two
concurrent send operations on both endpoints).
% The latter enforces two invariants, namely that no endpoint is
% used simultaneously by parallel processes and that the two endpoints of the same
% session have dual types.
The operational semantics is formulated as a standard reduction relation; hence,
proving subject reduction requires proving type preservation for structural
congruence.

@@

Given those peculiarities, the syntax and semantics of $\pi$-calculi have been
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and in so many encoding styles that it would take too long to
review.  They range from concrete encodings based on named syntax
(Melham's op.cit.) to basic de Bruijn (cite) and locally nameless
representation%~\cite{Chargueraud2012}
(\eg~\cite{Castro2020}).
Nominal approaches are also common, see~\cite{Bengtson2009}, but they
may be problematic w.r.t.~proof assistants based on dependent types.

@@

Some authors have already described the difficulties that combining techniques may induce in a mechanisation, e.g.~\cite{DBLP:conf/forte/ZalakainD21}.

>>>>

While results about concurrent
formalisms have already been mechanised (as we will discuss further
below), our experience is that choosing appropriate mechanisation
techniques and tools remains a significant challenge and that the
trade-offs are not well understood.  This often leads researchers
toward a trial-and-error approach, resulting in sub-optimal solutions,
wasted mechanisation effort, and techniques and results that are hard
to reuse. For example, the authors in~\cite{Cruz-Filipe2021b} note that the high-level parts of
mechanised proofs closely resemble the informal ones, while the
main challenge lies in getting the infrastructure right.

@@

\subsection{Linearity}
Linear typing systems allow  tracking the use of resources in a program. In
the case of typed process calculi (in particular, session types), linearity is
widely used for checking if and how a channel is used to send or receive values.
This leads to substructural typing~\cite[Chapter 1]{Pierce2004-oq} with
resulting mechanisation difficulties --- \eg deciding how to \emph{split the
typing context} in a parallel composition.

The goal of our challenge on linear reasoning is proving a type safety theorem
for a process calculus with session types, by combining as usual subject
reduction with absence of errors.  We have chosen  only to model linear (as
opposed to shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where a restriction $(\nu ab)\ldots$ binds two dual names $a$
and $b$ as opposite endpoints of a same channel --- and their duality is
reflected in the type system. We model a simple notion of error: a well-typed
processes must never use dual channel endpoints in a non-dual way (\eg by
performing concurrent send/receive operations on a same endpoint, or two
concurrent send operations on both endpoints).
The operational semantics is formulated as a standard reduction relation; hence,
proving subject reduction requires proving type preservation for structural
congruence.

@@

Given those peculiarities, the syntax and semantics of $\pi$-calculi have been
mechanised from an early age (see~\cite{Melham1994}) with many proof
assistants and in so many encoding styles that it would take too long to
review.  They range from concrete encodings based on named syntax
\cite{Melham1994} to basic de Bruijn \todo{cite} and locally nameless
representation%~\cite{Chargueraud2012}
(\eg~\cite{Castro2020}).
Nominal approaches are also common, see~\cite{Bengtson2009}, but they
may be problematic w.r.t.~proof assistants based on dependent types.

@@

% \cite{Brady2017} notes that using algebraic effects is unwieldy when there are many communication channels and suggests an experimental approach using dependent uniqueness types.
% \cite{Zalakain2019} notes that systems lacking linear types must simulate linearity, and that using a linearity predicate complicates congruences.
% They had to limit themselves to an asymmetric structural precongruence to be able to prove subject reduction.
Some authors have already described the difficulties that combining techniques may induce in a mechanisation, e.g.~\cite{DBLP:conf/forte/ZalakainD21}.
